---
title: 'Lab 1: Linear models for quantitative genetics'
author: "Abolfazl Arab - (or Abe)"
date: "10/16/2024"
output:
  html_document: default
  pdf_document: default
subtitle: BMI 206
---

```{R}
library(tidyverse)
```

### PART1: Analyzing provided genotype and phenotype data.

**Prepare the data.** Read in the genotype and phenotype matrices.

```{r}
genos = as.matrix(read.table("./genos.txt"))
phenos = as.matrix(read.table("./phenos.txt"))
```

Make a histogram of the phenotypes. Do they look normally distributed?

```{r}
hist(phenos)
```

*I think these looks like a normal distribution*.

<br>

How are the genotypes encoded?

```{r}
table(genos)
```

*grouped into "0", "1", and "2"*

<br>

How many individuals are there in the dataset and how many SNPs? (Save them in `N` and `M`, respectively.)

-   N = 1500 \# How many individuals

-   M = 10000 \# how many SNPs

```{r, eval=FALSE}
dim(genos)
dim(phenos)
N = 1500  # How many individuals 
M = 10000 # how many SNPs
```

<br>

**Compute the *minor* allele frequency for every SNP. Check MAFs are \<0.5.**

```{r, eval=FALSE}
MAFs = array(0,M)
for(i in 1:M) {
  ind = table(genos[,i]) %>% data.frame
  
  freq_A = sum( (ind$Freq / (2 * N) ) * c(0,1,2))

  MAFs[i] = min(c(freq_A, 1 - freq_A))

}
MAFs[1:10]
max(MAFs)
```

```{r, eval=FALSE}
MAFs %>% length
```

<br>

**Run a GWAS under an additive model and save the p-values, z-scores, and effect sizes.**

z-scores: estimate divided by standard-deviation

```{r, eval=FALSE}
pvalues = array(0,M)
zscores = array(0,M)
betas = array(0,M)

for(i in 1:M) {
	g = genos[,i]
	res = summary(lm(phenos~g))
	zscores[i] = res$coefficients[2,'t value']
	pvalues[i] = res$coefficients[2,'Pr(>|t|)']
	betas[i]   = res$coefficients[2,'Estimate']

}

```

<br>

Summarize the effect sizes.

```{r, eval=FALSE}
summary(betas)
hist(betas)
```

<br>

Are there any significantly associated SNPs? If so, which SNPs are they?

```{r, eval=FALSE}
assoc = which(pvalues<0.05 / length(pvalues))
assoc
```

<br>

How big are their effect sizes? How significant are they?

```{r, eval=FALSE}
betas[assoc] %>% min; betas[assoc] %>% max
zscores[assoc]
pvalues[assoc] %>% min; pvalues[assoc] %>% max
```

-   *effect sizes range from `-3.4` to `2.2`.*

-   *-log10(p-values) `07` to `86`*

<br>

Draw a QQ plot for log10(p) values.

```{r, eval=FALSE}
obsLogPvs = sort(-log10(pvalues))
expLogPvs = sort(-log10(seq(1/M,1,1/M)))
plot(expLogPvs,obsLogPvs,main='QQ plot')
abline( a=0, b=1 )
#label the significant SNPs red 
points(expLogPvs[(M-length(assoc)):M],obsLogPvs[(M-length(assoc)):M],col="red")
```

<br>

Is there inflation? Use the chi-square statistics to check.

```{r, eval=FALSE}
chis = zscores^2
lambdaGC = median(chis)/0.454 # why .454?
lambdaGC
```

*This `lambdaGC` suggests that there is some p-value inflation. I think the number of variants highlighted in red is biologically reasonable and we are not seeing huge number of significant data points.*

<br>

Plot the phenotype predictions for the most significant SNP.

```{r, eval=FALSE}
topSNP = genos[,order(pvalues)[1]]
plot(topSNP,phenos)
abline(lm(phenos~topSNP)$coeff,col="red")
```

<br>

**Build a linear predictor of the phenotype using the associated SNPs.**

```{r, eval=FALSE}
ypred = array(0,N)
for(i in 1:N) {
      ypred[i] = genos[i,assoc] %*% betas[assoc]
}

plot(ypred,phenos)
```

<br>

What is the correlation between the predicted phenotype and the true phenotype?

```{r, eval=FALSE}
cor(ypred,phenos)
```

<br>

**BONUS: Test each of the associated SNPs for non-linearity.**

```{r, eval=FALSE}
hp = array(0,length(assoc))

for (i in 1:length(assoc)) {
  g = genos[,assoc[i]]
  h = g
  h[h==2]=0
  #Hint: can use anova(lm(?),lm(?)) or summary(lm(?))
  hp[i] <- anova( lm(phenos~g), lm(phenos~g*h) )$Pr[2]
  #skip multiple test correction for now
}

hp
```

<br>

BONUS: Visualize a linear SNP and a non-linear SNP.

```{r, eval=FALSE}
par( mfrow=c(1,2) )


hp = array(0,M)

for (i in 1:M) {
  g = genos[,i]
  h = g
  h[h==2]=0
  #Hint: can use anova(lm(?),lm(?)) or summary(lm(?))
  hp[i] <- anova( lm(phenos~g), lm(phenos~g*h) )$Pr[2]
  #skip multiple test correction for now
}


linSNP = genos[,which.max(hp)]
nonlinSNP = genos[,which.min(hp)]

plot(linSNP,phenos)
points( c(0,1,2), tapply(phenos,linSNP, mean ), col=2, pch=16, cex=3 )
lines( c(0,1,2), tapply(phenos,linSNP, mean ), col=2, lwd=2  )

plot(nonlinSNP,phenos)
points( c(0,1,2), tapply( phenos,nonlinSNP, mean ), col=2, pch=16, cex=3 )
lines( c(0,1,2), tapply( phenos,nonlinSNP, mean ), col=2, lwd=2  )
```

<br>

**Repeat the GWAS to test for recessive rather than additive genetic effects.**

```{r, eval=FALSE}
genos2 = genos
genos2[genos<1]=1 # (AA)[1], (AG)[1], (GG)[2]
pvalues2 = array(0,M)
zscores2 = array(0,M)
betas2 = array(0,M)
for(i in 1:M) {
  g = genos2[,i]
  res = summary(lm(phenos~g))
  zscores2[i] = res$coefficients[2,'t value']
  pvalues2[i] = res$coefficients[2,'Pr(>|t|)']
  betas2[i]   = res$coefficients[2,'Estimate']
}
```

<br>

**Are the same SNPs significant or not?**

```{r, eval=FALSE}
assoc2 = which(pvalues2<0.05 / length(pvalues2))
assoc2
```

<br>

**How did the effect sizes change?**

```{r, eval=FALSE}
plot(betas,betas2)
```

```{r, eval=FALSE}
summary(betas)
summary(betas2)
```

*The scatter plot and summary stats show the difference!*

<br>

### PART2: Simulating genotypes with LD.

**Establish some important simulation parameters.**

```{r}
N = 1000 #number of individuals
M = 30   #number of non-causal SNPs
gs = matrix(0,nrow=N,ncol=M)
```

<br>

**Simulate a GWAS data set.** First, simulate the causal variant.

```{r}
set.seed = (42) #set random seed so we all get the same numbers
MAF = 0.5
gC = rbinom(N,1,MAF) #causal variant
```

<br>

Then, simulate the phenotypes given the causal variant.

```{r}
beta = 0.3 #association of causal variant
pheno = gC*beta + rnorm(N) 
```

<br>

Generate 10 SNPS in tight LD with the causal SNP.

```{r}
rho = 0.9

for(i in 1:10) {
  # idx: the chance they are going to be re-defined
  idx = rbinom(N,1,rho) 

  gs[,i]=gC*idx+rbinom(N,1,MAF)*(1-idx)
  
  
  # test they have the right LD empirically
  cat( 'Observed LD = ', cor( gs[,i], gC ), '\n' )
  # Bonus: prove they have the right LD theoretically
}
```

<br>

Do the same for 10 moderate LD partners (rho=0.6).

```{r,eval=FALSE}
rho = 0.6

for(i in 11:20) {
  idx = rbinom(N,1,rho) 

  gs[,i]=gC*idx+rbinom(N,1,MAF)*(1-idx)
  
  # test they have the right LD empirically
  cat( 'Observed LD = ', cor( gs[,i], gC ), '\n' )
  # Bonus: prove they have the right LD theoretically
}
```

<br>

Do the same for 10 independent SNPs (rho=0).

```{r,eval=FALSE}
rho = 0

for(i in 21:30) {
  idx = rbinom(N,1,rho) 

  gs[,i]=gC*idx+rbinom(N,1,MAF)*(1-idx)
  
  # test they have the right LD empirically
  cat( 'Observed LD = ', cor( gs[,i], gC ), '\n' )
  # Bonus: prove they have the right LD theoretically
}
```

**Run GWAS on the causal variant. Then run GWAS on the other variants. Keep track of the zscores only.**

```{r,eval=FALSE}
zsC = summary(lm(pheno~gC))$coef[2,3]
zs = sapply( 1:M, function(i) summary(lm(pheno~gs[,i]))$coef[2,3] )
```

<br>

Visualize the relationship between the mean z-scores at the tag SNPs and the z-score at the causal SNP.

```{r,eval=FALSE}
par( mfrow=c(2,2) )
breaks = hist(c(0,zsC,zs),plot=F)$breaks
hist(zs[1:10],breaks=breaks, col=1, main='LD partners')
abline(v=zsC)
hist(zs[11:20],breaks=breaks, col=2, main='Low-LD partner SNPs')
abline(v=zsC)
hist(zs[21:30],breaks=breaks, col=3, main='Independent SNPs')
abline(v=zsC)
```

<br>

**BONUS: Perform LD score regression. First, calculate the LD scores. There should be M+1 of them.**

```{r, eval=FALSE}
ldscores = ?
ldscores
```

<br>

BONUS: Visualize LD score regression.

```{r,eval=FALSE}
chis = c( ?, ? )^2
plot( ?, chis, ylab=expression(chi^2) )
#test for inflation
lambdaGC = median(chis)/0.454
lambdaGC
```

<br>

BONUS: Estimate heritability.

```{r,eval=FALSE}
summary( lm( ? )$coef[2,1] * M/N
```

<br>

BONUS: What is the true heritability?

```{r, eval=FALSE}
var(?) / var(?)
```
